---
title: "Problem 2 (Henry)"
date: "2026-10-09"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(readr)
library(dplyr)
library(stringr)
library(ggplot2)
library(tidyr)
```

# Problem 2

## Do students who have taken both 131 and 134 report greater interest in machine-learning-related fields than students who have taken 131 only?

We want to see if taking only PSTAT 131 has any correlation in machine learning domain interests.

```{r}
problem2 <- read_csv("~/Desktop/PSTAT197A/module0-class-survey-team9/data/problem2.csv")

dim(problem2)

head(problem2)
```

We are assuming that students who have taken PSTAT 134 have also taken PSTAT 131 since it is a prerequisite to that course. We confirmed this step by checking that every student who listed PSTAT 134 also included PSTAT 131.

```{r}
course_check <- problem2 %>%
  mutate(
    took_131 = str_detect(courses, "\\bPSTAT 131\\b"),
    took_134 = str_detect(courses, "\\bPSTAT 134\\b")
  )

# Show everyone who took PSTAT 134 but not PSTAT 131
course_check %>%
  filter(took_134 & !took_131) %>%
  select(response_id, courses)
```

Let's see the potential options for area interests:

```{r}
possible_values <- unique(unlist(strsplit(as.character(problem2$areas), split = ",")))

possible_values
```

Out of the eleven options, we will define 'machine learning' interests as one of the following six options:

```{r}
ml_areas <- c(
  "Analysis or classification of images",
  "Deep learning and neural networks",
  "Data acquisition and engineering",
  "Natural language processing and analysis of text",
  "Predictive modeling, generally",
  "Statistical models and inference, generally"
)
```

### PSTAT 131 vs non-PSTAT 131

```{r}
ml_data <- problem2 %>%
  mutate(
    took_131 = str_detect(
      coalesce(courses, ""),
      "\\bPSTAT 131\\b"
    ),
    areas = coalesce(areas, "")
  ) %>%
  separate_rows(areas, sep = ",\\s*") %>%
  mutate(selected = areas %in% ml_areas) %>%
  group_by(response_id, took_131) %>%
  summarise(
    n_ml_areas = sum(selected),
    .groups = "drop"
  ) %>%
  mutate(
    course_group = if_else(
      took_131,
      "Took PSTAT 131",
      "Did not take PSTAT 131"
    )
  )
```

```{r}
ml_data %>%
  group_by(course_group) %>%
  summarise(
    n = n(),
    mean_ml_areas = mean(n_ml_areas),
    median_ml_areas = median(n_ml_areas),
    .groups = "drop"
  )
```

```{r}
ggplot(ml_data, aes(x = course_group, y = n_ml_areas,
                    fill = course_group)) +
  geom_boxplot() +
  labs(
    title = "ML-Related Interests by PSTAT 131 Enrollment",
    x = "Course Group",
    y = "Number of ML-Related Areas Selected"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```

```{r}
wilcox.test(
  n_ml_areas ~ course_group,
  data = ml_data,
  exact = FALSE
)
```

### Both PSTAT 131 and PSTAT 134 vs Only PSTAT 131

```{r}
ml_data <- problem2 %>%
  mutate(
    took_131 = str_detect(courses, "\\bPSTAT 131\\b"),
    took_134 = str_detect(courses, "\\bPSTAT 134\\b"),
    course_group = case_when(
      took_131 & took_134 ~ "Both courses",
      took_131 & !took_134 ~ "PSTAT 131 only",
      TRUE ~ NA_character_
    ),
    areas = coalesce(areas, "")
  ) %>%
  filter(!is.na(course_group)) %>%
  separate_rows(areas, sep = ",\\s*") %>%
  mutate(selected = areas %in% ml_areas) %>%
  group_by(response_id, course_group) %>%
  summarise(
    n_ml_areas = sum(selected),
    .groups = "drop"
  )
```

```{r}
ml_data %>%
  group_by(course_group) %>%
  summarise(
    n = n(),
    mean_domains = mean(n_ml_areas),
    median_domains = median(n_ml_areas),
    .groups = "drop"
  )
```

```{r}
ggplot(ml_data, aes(x = course_group, y = n_ml_areas,
                    fill = course_group)) +
  geom_boxplot() +
  labs(
    title = "ML-Related Interests by Course-Taking Group",
    x = "Course-Taking Group",
    y = "Number of ML-Related Areas Selected"
  ) +
  theme_minimal() +
  theme(legend.position = "none")
```
